Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-03-28 rmd/manuscript.Rmd
edit: simulation update

State transitions shed light on how FCL changes occurred. In the chain motif, an additive mechanism (i.e., the addition of higher trophic species) was always responsible for the changes in FCL. With disturbance, state transition occurs from "B" to "B + C" (low productivity) or "B + C" to "B + C + P" (high productivity) as ecosystem size or complexity increases. Exceptionally, the state transition of "B + C + P" $\rightarrow$ "B + C" occurred as ecosystem complexity increased with no disturbance and low productivity, a scenario in which ecosystem complexity decreased FCL.

@@

High productivity strengthened the relationship between FCL and ecosystem geometry. In the chain motif, productivity greatly increased the overall FCL (**Figure S7**), thereby shifting the upper bound of FCL (**Figures 2 and 3**). In contrast, increased productivity greatly decreased the minimum levels of FCL with weak and strong omnivory with little influence on the maximum FCL (**Figures 2, 3, and S7**). This negative effect of increased productivity was apparent in smaller and/or simpler ecosystems, steepening the slopes between FCL and ecosystem geometry (**Figures 2 and 3**).

@@

The ecosystem size hypothesis of FCL has received support from both theoretical and empirical research [@vanderzandenPatternsFoodChain1999; @postEcosystemSizeDetermines2000; @takimotoEffectsProductivityDisturbance2012], although empirical work so far has rarely elaborated on the underlying drivers [but see @saboRoleDischargeVariation2010; @wardMechanisticTheoryAquatic2017]. In contrast, ecosystem complexity has not been incorporated in theoretical explorations of FCL to date. Here, we show that FCL responds positively to increasing ecosystem size and complexity, assessed as the number of patches and branching probability in river networks. This supports our hypothesis that ecosystem complexity may represent an overlooked yet important extrinsic control of FCL. However, the relationship between FCL and ecosystem complexity varied with other extrinsic and intrinsic factors, supporting the emerging paradigm that FCL is regulated by multiple factors [@mchughDualInfluencesEcosystem2010; @wardMechanisticTheoryAquatic2017]. Although we focus on river networks here, we expect our results to be transferable to other systems with non-branching structures. Ecosystems are intrinsically complex -- forests possess complex vertical and horizontal structures, which may provide heterogeneous microhabitats with different levels of vulnerability to wind and/or fire disturbances [@canslerClimateFireSize2014; @petersonContagiousDisturbanceEcological2002; @rommeHistoricalPerspectiveYellowstone1989]. Topographic and terrain variation can also mediate spatial disturbance patterns in a landscape [@connell30YearStudy1997; @stueveSpatialPatternsIce2007]. Explicit consideration of ecosystem complexity may therefore increase our ability to predict FCL and complements spatial food web theory that has emerged over the past two decades [@holtFoodWebsSpace2002a; @mccannDynamicsSpatiallyCoupled2005; @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012; @andersonEffectsDispersalRiver2018; @teruiSpatialDisturbanceSynchrony2019].

>>>>

State transitions shed light on how FCL changes occurred. In the chain motif, an additive mechanism (i.e., the addition of higher trophic species) was always responsible for the changes in FCL. With disturbance, state transition occurs from "B" to "B + C" (low productivity) or "B + C" to "B + C + P" (high productivity) as ecosystem size or complexity increases (**Figures S5 and S6**). Exceptionally, the state transition of "B + C + P" $\rightarrow$ "B + C" occurred as ecosystem complexity increased with no disturbance and low productivity, a scenario in which ecosystem complexity decreased FCL.

@@

High productivity strengthened the relationship between FCL and ecosystem geometry. In the chain motif, productivity greatly increased the overall FCL (**Figure S7**), thereby shifting the upper bound of FCL (**Figures 2 and 3**). In contrast, increased productivity decreased FCL with weak and strong omnivory (**Figures 2, 3, and S7**). This negative effect of increased productivity was apparent in smaller and/or simpler ecosystems, steepening the slopes between FCL and ecosystem geometry (**Figures 2 and 3**).

@@

The ecosystem size hypothesis of FCL has received support from both theoretical and empirical research [@vanderzandenPatternsFoodChain1999; @postEcosystemSizeDetermines2000; @takimotoEffectsProductivityDisturbance2012], although empirical work so far has rarely elaborated on the underlying drivers [but see @saboRoleDischargeVariation2010; @wardMechanisticTheoryAquatic2017]. In contrast, ecosystem complexity has not been incorporated in theoretical explorations of FCL to date. Here, we show that FCL responds positively to increasing ecosystem size and complexity, assessed as the number of patches and branching probability in river networks. This supports our hypothesis that ecosystem complexity may represent an overlooked yet important extrinsic control of FCL. However, the relationship between FCL and ecosystem complexity varied with other extrinsic and intrinsic factors, supporting the emerging paradigm that FCL is regulated by multiple factors [@mchughDualInfluencesEcosystem2010; @wardMechanisticTheoryAquatic2017]. Although we focus on river networks here, we expect our results to be transferable to other systems with non-branching structures. Ecosystems are intrinsically complex -- forests possess complex vertical and horizontal structures, which may provide heterogeneous microhabitats with different levels of vulnerability to wind and/or fire disturbances [@canslerClimateFireSize2014; @petersonContagiousDisturbanceEcological2002; @rommeHistoricalPerspectiveYellowstone1989]. Topographic and terrain variation can also mediate spatial disturbance patterns in a landscape [@connell30YearStudy1997; @stueveSpatialPatternsIce2007]. Explicit consideration of ecosystem complexity may therefore increase our ability to predict FCL and complements spatial food web theory that has emerged over the past two decades [@holtFoodWebsSpace2002; @mccannDynamicsSpatiallyCoupled2005; @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012; @andersonEffectsDispersalRiver2018; @teruiSpatialDisturbanceSynchrony2019].
